---
title: "Quick R/Python Beginner Guide for Ecologists"
author: "Oleksii Dubovyk"
site: bookdown::bookdown_site
output: bookdown::bs4_book
mainfont: Tempora
fontsize: 11pt
output_dir: "docs"
---

# About{-}

This is a quick guide to the basics of programming with R and Python. Treat it is a cheatsheet. Maybe I will get enough time to extend it into a proper guide.

# Getting started (what is IDE?)

You can use any programming language in a command line (on Windows, call `Win+R`, write "cmd" and hit Enter, on Mac -- open app called "Terminal", on Linux -- hit `Ctrl+Alt+T`). The command line is the most primitive way to interact with your computer through short statements: you can create or delete new directories, files, list folder contents, and much-much more. Although you theoretically can write long and complex commands, your machine will only interpret a single line at a time. So doing anything but simple tasks using a command line is not very convenient.

With any programming language, you might want to have your program contained in a script. This way, instead of typing a set of commands in the command line, you can edit and save the entire script and run it as a file.

The most convenient way to deal with scripts is using an IDE -- integrated development environment. IDE is just another computer program that allows you to open, edit, save, and run scripts in your programming language of choice. Ideally, an IDE should work with any language (although for many of them there is also a compiling step in running a code), but surely enough some are more popular than the others.

## Python

### Installing Python

To use a language, you need to install that language on your machine. You will need to do some digging to install a language, and for Python, here is a good starting point: [https://www.python.org/downloads/](https://www.python.org/downloads/).

Ubuntu users can easily type in Terminal:

```{bash eval=FALSE}
sudo apt-get update
sudo apt-get install python3
```

### Visual Studio Code

Visual Studio Code is one of the IDEs that is supposed to be used for any language. I don't use it often though. You can download it here: [https://code.visualstudio.com/download](https://code.visualstudio.com/download).

### PyCharm

PyCharm is pretty simple and lies somewhere between a text editor and a sophisticated IDE. It can be pretty useful to edit an existing code, but might be not enough to write a sophisticated script from scratch. You can download it here: [https://www.jetbrains.com/pycharm/download/](https://www.jetbrains.com/pycharm/download/).

### Spyder

Spyder is probably one of the most complex IDEs, which will be especially intuitive for the past R/RStudio users. However, if you decide to use it, you might need to be at the level of experience where you know what `pip` or `anaconda` is, and dealing with those can be extremely annoying. Here is the download link: [https://docs.spyder-ide.org/current/installation.html](https://docs.spyder-ide.org/current/installation.html).

### RStudio

Yes, that's right, RStudio for Python. The current philosophy of RStudio developers is to create a cross-platform IDE (that's why they are renaming RStudio to Posit). We will get to RStudio later.

### Jupiter

Jupiter notebooks are pretty common nowadays because they allow you to write a whole interactive document (e.g., report, paper). This means that instead of a dry Python code, you write a real document, like a notebook, where you can embed chunks of code that will run and do something. A Jupiter notebook is something like what this guide looks like -- mostly I write text, but hey, there are boxes of code that you can copy and run.

A Jupiter notebook could be a good starting point for somebody who is learning Python. For example, if you don't want to install Python locally on your computer, you can use a cloud-based [Colab](https://colab.research.google.com/). The disadvantage is that all your code will be run somewhere on Google servers, so there might be a delay; besides, you need a stable Internet connection.

## R

### Installing R

To install R, just follow the instructions: [https://cran.r-project.org/](https://cran.r-project.org/).

### RStudio

There are other IDEs for R, but RStudio is the most prevalent one. You can download it here: [https://posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/).

Note: RStudio and R are two different things. In your CV, you might want to say that you know R (the language), but don't say that you know RStudio (the tool) -- that serves as a red flag that you don't really know what you are talking about. Saying "I have skills with RStudio" is like saying "I can use a keyboard" instead of "I am proficient in English".

# Objects and data types

Both Python and R are object-oriented high-level languages. The term "high-level" means that the programming language is written in some lower-level programming languages (e.g., C++), and thus are made to be more understandable by a human programmer (e.g., you don't need to think about how much RAM you need to allocate to create and interact with an object -- the language does it for you behind the scenes).

The term "object-oriented" means that the definition of objects is central in this programming. An object is any piece of data: from a single value to entire complex datasets. In object-oriented programming, objects are often assigned to certain classes of objects. Each object from a class have certain attribute and methods that can be applied to it.

For example, I have a succulent plant. I call it Steve. If it was an object in my computer, we could define the following:

- class: plant

- object: Steve

- attributes: age, number of leaves, biomass, etc.

- methods: water, plant in a larger pot, add nutrients.

Python and R have slightly different approaches to operating objects, but the logic is pretty similar. But before we get there, let's talk about the simplest types of data that can be stored in an object.

## Python

- Boolean: the simplest logical data type, a single value out of two possible values; a flip of a coin would be a good example, and the possible values can be viewed as "heads" or "tails", "yes" or "no", "1" or "0", "True" or "False"; in Python, the Boolean values are denoted as `True` or `False`;

- Numeric: this would be a number, and there are actually different types of a number:

  - Integer: this is any real whole number: `1`, `5000`, `-10000000`, etc.
  
  - Float: any non-integer number, e.g., `1.235`,
  
  - Complex number: you wouldn't use it unless you are deep into math
  
- Sequence: any ordered sequence of values

  - String: a sequence of characters, in Python defined with single or double quotation marks, e.g., `'Hello world'` or `"Hello world"`; the length of a string can be any positive integer, from a single character to a whole paragraph of text
  
  - List: an ordered list of objects of any type denoted with square brackets, e.g., `[1, True, 0.25, "apple"]`; lists can be modified after they are created; a list element can also be a list, e.g., `[[1, True, 0.25, "apple"], "second element here"]`
  
  - Tuple: pretty much the same as a list, but cannot be modified (is immutable); a tuple is defined with parentheses, e.g., `(1, True, 0.25)`
  
- Dictionary: a set of key-value pairs like you would have in a dictionary, e.g., key "one"-value "uno", defined with curly brackets as `{"key 1": "value 1", "key 2": "value 2"}`

- Set: an unordered set of unique values defined with a function `set()`

## R

- Logical: same as Boolean in Python, defined as `TRUE` or `FALSE` that also can be abbreviated as `T` or `F`

- Numeric: any non-complex number;

- Character: same as string in Python, e.g., `"Hello world"`

Some more complex data structures in R include vectors, data.frames, factors, which will be discussed later.

# Basic syntax

## Python

### Create and play with variables

Say, you want to create add two numbers. Sure, you definitely can do that:

```{python}
2 + 3
```

This all is not too useful though. We are not using the perks of objects, and the statement above is not much better than having a calculator. Let's create two objects and assign integers to them.

```{python}
a = 2
b = 3
c = a + b
print(c)
```

The function `print()` here prints out a value of an object. Without that last line, all the operation would have still happened but we would never see the result. Sowe can see that we are operating on these objects. Let's check what class the object `c` belong to. The function `type()` allows us to look up the class of an object:

```{python}
type(c)
```

There are not that many operations we can do with an integer. Let's rewrite one of them into a float.

```{python}
b = 3.5
c = a + b
print(c)
```

Now we can apply different functions to this object, for example, round it up, which returns an integer.

```{python}
round(c)
type(round(c))
```

You can name your objects with almost any name, except for the reserved ones (e.g., you can't name an object `True`, because `True` is a Boolean value). Let's see how to create objects of other classes.

```{python}
x1 = False
```

Let's check if `x1` is False:

```{python}
x1 == False
```

So it is true that the object `x1` is `False`.

What about strings?

```{python}
x2 = "False"
x2 == False
```

It is not true now, because `x2` is nothing but a string of characters and not a Boolean variable.

You can index a string, i.e., access a specific element (character) in a string: just like "H" is the first letter in "Hello", "e" is the second, and so on. Be aware, however, that indexing in Python starts with zero, so that the first element has an index [0], the second -- [1], the third -- [2], and so on. You can also index a sequence from the end using a minus sign, so that the last character is indexed as [-1], the second to the last -- [-2]. You can also specify a range of indexed values with a colon, so that [0:3] denotes characters from the first to the fourth. Indexing in Python is specified with square brackets:

```{python}
x2
x2[0]
x2[1]
x2[2]
x2[-1]
x2[0:3]
```

In Python, the plus operator applied to strings concatenates them:

```{python}
x2[0:3] + x2[1] + x2[0] + x2[4] + x2[2]
```

Similar rules apply to lists. You can index lists, e.g.,

```{python}
x3 = list(x2)
x3
x3[0:3]
```

Lists have a certain number of methods. One of the common ones is appending with another object. Notice how methods in Python are applied to objects with a period, so that you have `object.method(argument)`.

```{python}
x4 = [1, 2, 3, 4]
x4
x4.append(5)
x4
```

Each class of objects has a specific list of methods that can be applied to such objects. You might need to read python documentation for different classes (e.g., [here](https://docs.python.org/3/tutorial/datastructures.html)). You also can save an output of a method into another object:

```{python}
x4.append(5)
x5 = x4.count(5)
x5
```

### Operators

Objects in programming languages are often handled with certain operators. Here are the most common ones:

- Assignment:

  - `=` -- assign
  
  - `+=` -- add and assign (`x += 1` is the same as `x = x + 1`)
  
- Comparison:

  - `==` -- check if equal (output is either `True` or `False`)

  - `!=` -- check if not equal

  - `>`, `>=`, `<`, `<=` -- check if larger than, larger or equal to, less than, less or equal to

- Arithmetic:

  - `+`, `-`, `*`, `/` -- add, subtract, multiply, divide

  - `**` -- exponentiation
  
  - `%` -- modulo (remainder from division)

  - `//` -- floor division (integer of a division)

- Logical:

  - `and`, `or`, `not` -- used in Boolean expressions, e.g., to check if both expressions are satisfied (`and`), or either is satisfied (`or`)
  
  - `is`, `is not` -- check whether objects are identical
  
- Membership:

  - `in`, `not in` -- checks if an element is a member of some list or other sequence

### Comments

It may be useful to leave some verbal comments in your code that wouldn't be read and executed by the computer. In Python, hash tag comments out a line of code or everything after it:

```{r}
a = 2
b = 3
# b = 5000
a + b # still 5, so b was never assigned as 5000. just like this text
```

### Loops

Loops are used for iterations over a sequence, meaning that you go over each element and do something with it. `for` loops are more common and do something *for* each element, and `while` loops conduct iterations as long as some logical condition is satisfied. For example,

```{python}
for i in x4:
  print(i)
```

does pretty much the following: for each element of `x4`, let's call this element `i`, and then print this `i`.

In the `while` loop, on the other hand

```{python}
number = 1 # initialize the variable
# you can explicitly define the class, in which case it would be
# number = int(1)

while number < 5:
  print(number)
  number += 1 # same as "number = number + 1"
```

which the computer understands as the following: create a variable `number` and assign it a value of `1`; then run an iteration, at each step of which print out the value of the `number` and increase its value by one; this process is repeated as long as the statement `number < 5` is `True`; once `number` becomes `5`, this0 statement becomes `False` and the loop stops.

While loops are a bit tricky because you can accidentally end up with an infinite loop. For example,

```{python eval=F}
while True:
  print("AAAAAA")
```

If you run it, your computer will never stop yelling at you because `True` is always `True`, and the while loop execute every next iteration.

### If/else

What if you run a loop and want different outcomes under different conditions? That's where you would use an `if`/`else` statement. For example, you for your sequence you can check if a number is odd or even:

```{python}
for i in x4:
  if i % 2 == 0:
    print(str(i)+" is even");
  else:
    print(str(i)+" is odd")
```

Here, we apply a modulo operator `%` to our value `i` to check the reminder of dividing `i/2` equals zero; if `True`, then `i` is even, else it is odd.

There could be more complex logical operations. What is the sequence represents something that we want to classify into three levels: low (0--1), medium (2--4), or high (5 and higher):

```{python}
for i in x4:
  if i <= 1:
    print(str(i)+" is low");
  elif i > 1 and i <= 4:
    print(str(i)+" is medium");
  else:
    print(str(i)+" is high");
```

where `elif` stands for a short "else if".

### Functions

Just like in math, a function is something that receives some input, or argument, and returns some input. For example, we want to write a function $f(x) = x^2$:

```{python}
def f(x):
  return x**2
```

`def` means "define a function", which the computer will memorize as `f`, it has an argument `x`, and the function exponentiates the argument to the power of 2. You can write an output of a function into a different variable.

```{python}
y = f(2)
y
```

You can apply functions in a loop, e.g.,

```{python}
for i in x4:
  print(f(i))
```

A function can have several arguments. For example, what if we want to use power different from 2?

```{python}
def exp(x, e = 2):
  return x**e
```

In that function, we defined a default option for argument `e`. Now, if the function doesn't receive the value of this argument, it will always assume that `e = 2`.

It is a good idea to define the arguments, but otherwise Python will read them in the order they were written in the function. So that the code `exp(2, 5)` will generate the same output as `exp(x = 2, e = 5)`.

Note that loops and Python in general are indent-sensitive: all statements in a loop or function should be indented in a stepwise manner. For example, this code will throw out an error because there is an indent missing:

```{python error=T}
def exp(x, e = 2):
return x**e
```


### Libraries

Sometimes you want to do something, but there is no tool for it in the basic syntax and it is too hard to program on your own. Well, chances are, some programmers have already done that, packed their code into an archive somewhere online, and you can easily reuse their functions. Those archives are often called "packages", "libraries", or "modules". To use a module, you need to download and build it on your machine (which usually is done with designated software), and load it into your memory before you use it. This way, you won't need to re-type somebody else's code in your script (because it can be thousands of lines long).

For example, I want to generate a random number. Though it seems simple, there is some math involved and I don't want to waste time tackling this problem. Well there exists a Python module for random number generation called `random``.

Here is the part I don't like too much compared to R: installing a module can be tricky, especially for a beginner. And if for some reason you need several modules that are interdependent, get ready to spend a day solving compatibility issues between them.

Anyway, you might need to [install](https://pip.pypa.io/en/stable/installation/) `pip` on your computer. This little guy, a package manager, is supposed to make it easier to install modules and avoid dependency errors. After installed, you should run pip through the command line like this (erm, it actually won't work because `random` is comes with Python when you install the language, but this is the syntax you could use when installing a new package):

```{bash eval=F}
pip install random
```

Now, at the beginning of a script, we need to tell Python the module to load. After that is done, you can call objects from the module like `module.object`:

```{python}
import random # load the module

random_list = [] # initialize the list

for i in range(10): # 10 iterations
  x = random.randint(0, 100) # create an intermediate random integer value between 0 and 11
  random_list.append(x) # append the list

print(random_list)
```

You can also use aliases in module names and their parts as following:

```{python}
import random as rand

random_list = []

for i in range(10):
  x = rand.randint(0, 100)
  random_list.append(x)

print(random_list)
```

### Lambda functions

Lambda functions are very simple and short syntax features that are useful for tasks where you don't really need to write a whole function. Once you run a lambda function, it is deleted from your memory. 

```{python}
def big_function(a, b, add = True):
  if add:
    small_function = lambda a, b : a + b
  else:
    small_function = lambda a, b : a - b
  output = small_function(a, b)
  return output
```

In this function, we first check if the argument `add` is True (note that `if add` and `if add == True` will produce the same result since if `add = True`, then the statement `add == True` will produce `True`). Depending on it, we create a `small_function` which adds or subtracts the arguments. Then we just apply the `small_function` to the `big_function`'s arguments `a` and `b` without a need to define clumsy `small_function` with `def` and all that stuff.

### List comprehension

List comprehension is an interesting technique that could be used to create sequences that follow a certain pattern without writing a clumsy `for` loop. For example, the statement

```{python}
doubles1 = [x*2 for x in range(0, 10)]
```

produces the same output as

```{python}
doubles2 = []

for x in range(0, 10):
  doubles2.append(x*2);
```

Let's compare them:

```{python}
doubles1
doubles2
```

You can also embed `if` statements in comprehended lists, e.g.,

```{python}
[x**2 for x in range(0, 10) if x>3]
```


## R

### Create and play with variables

Python style of assigning values to objects with an equal sign (e.g., `x = 1`) is valid in R. However, it is far more popular (and easier to read) to use an "arrow" operator `<-`, which in RStudio can be done through a shortcut LAlt + - (left Alt and minus sign).

```{r}
x <- 1
x
```

Boolean values are written as `TRUE`/`FALSE` or abbreviated `T`/`F`:

```{r}
FALSE == F
```

So it is true that false is false.

Characters and numeric objects are created in the same way,

```{r}
x1 <- 1.5
x2 <- "x2"
```

Sequences of objects are called vectors and are created using a function `c()`

```{r}
v1 <- c(1, 2, 3, 4, 5)
v1
```

You can also create sequences of integers using colon, 

```{r}
1:10
```

Or sequences of values using dedicated functions

```{r}
seq(1, 5, 0.5)
seq(from = 1, to = 5, length.out = 4)
```

Note that indexing in R, unlike Python, begins with 1:

```{r}
v1[1]
v1[1:3]
v1[-4]
```

Vectors can only contain objects of the same class. R will either throw an error or try to change all objects to one class if you try to violate this rule:

```{r}
c(1, 2, "three")
```

The simplest combination of vectors is a matrix. It is the same as a matrix in math and you can do linear algebra in basic R.

```{r}
matrix(data = 1:9, ncol = 3, nrow = 3, byrow = T)
```


Whatever is called lists in R is different from Python. It is more similar to JSON or XML kind of objects, and perhaps closer to Python dictionaries. It can be used to store different types of data with different dimensionality.

```{r}
l1 <- list("first element" = T,
           "second element" = c(1, 2, 3, 4),
           "third element" = matrix(data = 1:9, ncol = 3, nrow = 3))
l1
```

Note that R is not an indent-sensitive language and you can hit the next line for better readability. It all gets executed as a single line:

```{r}
l2 <- list("first element" = T, "second element" = c(1, 2, 3, 4), "third element" = matrix(data = 1:9, ncol = 3, nrow = 3))
l2
```

You can index list elements by their index or name, e.g., (note that you have to use double brackets to index list elements)

```{r}
l1[["first element"]]
l1[[3]]
l1[[2]][3]
```

### Operators and comments

There are some differences in notation compared to the operators used in Python.

Note that R operators work on vectors. In all fairness, what I said earlier about R object types wasn't exactly correct: all single-element objects are vectors, so it's not like "a vector is a sequence of several (say) numeric objects", but "a single numeric value is just a numeric vector of length 1". So look what happens when a simple arithmetic operator is applied to vectors:

```{r}
(-2:2) + (1:5)
```

The operation is repeated for all elements of the vector.

- Assignment:

  - `=` or `<-` -- assign
  
  - there is also a right assignment `->` but I have never used it
  
- Comparison:

  - `==` -- check if equal (output is either `True` or `False`)

  - `!=` -- check if not equal

  - `>`, `>=`, `<`, `<=` -- check if larger than, larger or equal to, less than, less or equal to

- Arithmetic:

  - `+`, `-`, `*`, `/` -- add, subtract, multiply, divide

  - `^` or `**` -- exponentiation
  
  - `%%` -- modulo (remainder from division)

  - `%/%` -- floor division (integer of a division)
  
  - `%*%` -- matrix multiplication

- Logical:

  - `&` -- vectorized AND operator
  
  - `&&` -- elemental AND operator, should only be used with single-element statements,
  
  - `|` and `||` -- vectorized and elemental OR operator, see below what the difference is
  
- Membership:

  - `%in%` -- checks if an element is a member of a vector; there is no "not in" operator so I often define it in my scripts: `'%!in%' <- function(x,y)!('%in%'(x,y))`
  
- Pipe:

  - `%>%` if you use tidyverse or `|>` in latest R distributions: the idea of pipe to make a statement where you apply a result of one function to another function more readable, so instead of code that reads `function4(function3(function2(function1(x))))` you could write
  
```{r eval=F}
x %>%
  function1() %>%
  function2() %>%
  function3() %>%
  function4()
```

Regarding the difference between vectorized and elemental logical operators, vectorized ones always apply to all elements of a vector, e.g.,

```{r}
((-2:2) >= 0) & ((-2:2) <= 0)
```

First, it evaluates the statements `((-2:2) >= 0)` (outputs `F F T T T`) and `((-2:2) <= 0)` (outputs `T T T F F`), and then applies the `&` operator to all five pairs, yielding only one that is `T` and `T` in both vectors.

Now, the statement `((-2:2) >= 0) && ((-2:2) <= 0)` will throw an error because `&&` can only be used with two single-element objects. It can be useful when you must be sure that you check single logical values and if a vector creeps in your code, then this code won't run.

As far as the comments, it's same as in Python, everything between a `#` sign and the end of the line is commented out and is not executed by R.

### Loops and if/else

Syntax is fairly similar to Python, but again, indentation is not important in R. Unlike Python, you are expected to separate the statement you iterate over with parentheses, while the iterative statement is contained within curly brackets. 

```{r}
odds <- numeric() # initialize empty numeric vector
evens <- numeric()

for (i in 1:10){
  if (i %% 2 == 0){
    odds <- c(odds, i) # there is no append command so you have to rewrite the object at each step
  }else{
    evens <- c(evens, i)
  }
}

odds
evens
```

A neat function in R that makes if/else functions easier is `ifelse` (just like in Excel), which lets you fit a whole statement in one line:

```{r}
x <- 1
ifelse(x == 1, "x equals 1", "x isn't equal to 1")
```

### Functions

Functions in R are defined as following:

```{r}
e <- function(x, e = 2){
  return(x^e)
}
```

So it's very similar to Python. You don't even need to explicitly say what to return, just a statement would suffice, although typing `return` is a good habit: sometimes your function can analyze a huge dataset and return nothing because you didn't explicitly wrote it to return anything! Also, with short functions, you don't even need to use any curly brackets, so the function above can be stripped all the way to

```{r}
e <- function(x, e = 2) x^e
```

### Apply and such

`apply` is a family of functions that can be used to iterate a function over a multi-element object, just like list comprehension in Python (although apply can be much more complex). For example, 

```{r}
sapply(1:5, e)
```

We just defined the function `e` above so that `apply` knows what it is, but more often you would write temporary functions directly in there (just like lambda functions in Python):

```{r}
sapply(1:5, function(x){
  if (x %% 2 == 0){
    paste(x, "is even") # btw, + doesn't concatenate strings in R like in python
  }else{
    paste(x, "is odd")
  }
})
```

`sapply` is used for simple vectors, `lapply` -- to iterate over a list, and `apply` -- to two-dimensional datasets where you also have to specify a dimension (`1` for row-wise operations, `2` -- for column-vise), e.g.,

```{r}
x <- matrix(1:9, nrow = 3, ncol = 3)
x

apply(x, 1, sum)
apply(x, 2, sum)
```


### Libraries

R libraries are the same as modules in Python. However, there is a [central archive called CRAN](https://cran.r-project.org/) for R libraries that is maintained by the same people who develop and maintain the R language. So you don't need any external `pip`s and such, just type directly in R language `install.packages("YOUR-PACKAGE-NAME")` and it will get installed on your computer. At the beginning of each session, you have to call the package you want to use with a statement `library(yourpackagename)`. Note that we use quotes for `install.packages` but not in `library`.

CRAN has quite strict requirements to packages, and some packages can be found elsewhere, e.g., GitHub. For example, in my undergrad I developed some utilities for my work that really are redundant relative to big packages like `vegan`, that is very popular in ecology. You still can install such packages using another package, `devtools`:

```{r eval=F}
install.packages("devtools") # remember that you only call this once to install it on your computer
library(devtools) # but this one you call every time you start a new R session
install_github("OleksiiDubovyk/ancends") # also only call once
library(ancends) # and here is my baby
```

